Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-15 examples.tex
stuff

One of the key tasks of a page fault handler in a general-purpose OS kernel is to map new pages into an address space by writing into an existing page table. To do so, we first allocate a fresh page, then calculate the appropriate known-valid page table walks and update the appropriate L1 page table entry.

@@

$\specline{\exists \texttt{ fpaddr} \ldotp \ulcorner \texttt{aligned fpaddr} \urcorner \ast \texttt{rax}\mapsto_{\textsf{r}} \texttt{fpaddr} \ast \texttt{fpaddr} \mapsto_{\textsf{p}} (\texttt{wzero 64})  \ulcorner \texttt{entry\_present fpaddr}\urcorner}_{\rtv}$

@@

$\specline{\exists \texttt{ fpaddr} \ldotp \ulcorner \texttt{aligned fpaddr} \urcorner \ast \texttt{rax}\mapsto_{r} \texttt{fpaddr} \ast \texttt{fpaddr} \mapsto_{p} (\texttt{wzero 64})  \ulcorner \texttt{entry\_present fpaddr}\urcorner}_{\rtv}$

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace} (<[\vaddr:=\texttt{fpaddr}]> \theta,m) \ast \vaddr \mapsto_{\textsf{vpte}}\; \qfrac \;\fpaddr \textsf{ wzero 64}}_{\rtv}$

@@

  \todo[inline,color=yellow]{Colin, could you speak about the fact that we have wpd in iProp so that's the reason to have rtval whereever table traversal is needed}

@@

\todo[inline]{this is terribly ugly, should switch to minted}

@@

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh page backing this virtual first requires an assurance on the existence L1 table entry which is realized in two folds:

@@

\item physical pointsto assertions on the tables L4, L3, and L2 reaching to the L1 level entry (l1e) ensures the successful table lookups with the fact on \textit{presence} (\textsf{entry\_present}) of the relevant entries (\textsf{l4e l3e l2e}) (Specification Lines 4-6)
\item ,and a virtual L1 pointsto ($\mapsto_{\textsf{vpte}}$) on the address computed from the reached L1 entry ($ \ulcorner \texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$) which is almost exactly same with the virtual-pointsto definition in Figure \ref{fig:virtualpointstosharing} except the physical page address is already computed, not exsitentially quantified:

@@

    \lambda \mathit{cr3val}\ldotp

@@

    \sumapaces{\mathit{cr3val}}\delta \ast 

@@

@@

After obtaining an virtual address \textsf{pte\_addr} in \textsf{rax} referring to a physical address, computed from already known L4-L2 physical memory pointstos ($\texttt{addr\_L1}(\vaddr,\entryo) = \paddr$),
in Specification Line 7 ($\texttt{pte\_addr} \mapsto_{\textsf{vpte}}  \; \paddr\;(\texttt{wzero 64})$), we save it to \textsf{r14} to be updated later in Line 9.

Then we allocate a fresh page initialized (i.e. zeroed) in Line 14 which return an \textsf{aligned} address of fresh page (\textsf{fpaddr}) in \textsf{rax} (Specification Line 19).

The crux point in the proof is updating L1 table through a virtual address (\textsf{pte\_addr}) through making a known physical address, in our example L1 table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$), to show the fresly allocated page address (\textsf{fpaddr}) in Line 22 to obtain
\[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by Rule} \TirNameStyle{WriteToVirtMemFromReg})\]
by Rule \TirNameStyle{WriteToVirtMemFromReg} (Specification Line 26). 
Consequently, right at this point in the proof, when we unfold the $\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr})$ we will be able obtain a full ownerhsip on the L1 table physical pointsto $\paddr \mapsto_{\mathsf{p}} \textsf{fpaddr}$ . Then, we can first discard the allocated virtual-address pointsto assertion (i.e. dropping \textsf{pte\_addr}), and split the full-ownership of physical-page pointsto ($\textsf{fpaddr}\mapsto_{\textsf{p}} \;\textsf{wzero 64}$) to obtain a fractional ownership (q4) on it. Since we know $\texttt{addr\_L1 }(\vaddr, \entryo) \mapsto_{\mathsf{p}} \{q4\} \;\textsf{fpaddr}$) from $\mapsto_{\textsf{vpte}}$, after the drop of virtual-pointsto, we still can connect L4-L2 pointsto assertions (in Specification Lines 24 and 25) with newly allocated $\texttt{fpaddr} \mapsto_{a} (\texttt{wzero 64})$ to obtain the complete physical table-walk assertion (L4\_L1\_PointsTo(maddr l4e l3e l2e fpaddr)) in Line 30.

Finally, we can insert $\vaddr$ into the ghost page-table-walk summarization map ($\theta$) to change its state from unmapped ($\ulcorner \theta \;!!\;\vaddr = \texttt{None}\urcorner$ in Specification Line 30) to mapped (Specification Line 33) by Iris' ghost-map update ($\sqsubseteq$), and construct virtual-pointsto $\vaddr \mapsto_{\textsf{vpte}}\; \qfrac \;\fpaddr \textsf{ wzero 64}$ (Specification Line 36) with $\sumwalkabs\vaddr\qfrac\fpaddr$ obtained from ghost page-table-walk insertions, $\sumapaces\rtv\delta$ from unfolding the definition of $\mapsto_{\textsf{vpte}}$ in Specification Line 18, and $\fpaddr \mapsto_{\textsf{p}} \textsf{wzero 64}$ in Specification Line 34. fragmental ownership of 
\subsection{Unmapping a Page}
The reverse operation, unmapping a designated page that is currently mapped, is pleasingly the reverse: by flipping the entry back to an invalid entry, full ownership of the updated entry reverts back to the now-invalid cell of the L1 table.

>>>>

One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table.
To do so, we first allocate a fresh page, then calculate the appropriate
known-valid page table walks and update the appropriate L1 page table entry.

@@

$\specline{\exists \texttt{ fpaddr} \ldotp \ulcorner \texttt{aligned fpaddr} \urcorner \ast \texttt{rax}\mapsto_{\textsf{r}} \texttt{fpaddr} \ast \texttt{fpaddr} \mapsto_{\textsf{p}} (\texttt{wzero 64}) \ast \ulcorner \texttt{entry\_present fpaddr}\urcorner}_{\rtv}$

@@

$\specline{\exists \texttt{ fpaddr} \ldotp \ulcorner \texttt{aligned fpaddr} \urcorner \ast \texttt{rax}\mapsto_{r} \texttt{fpaddr} \ast \texttt{fpaddr} \mapsto_{p} (\texttt{wzero 64}) \ast \ulcorner \texttt{entry\_present fpaddr}\urcorner}_{\rtv}$

@@

  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace} (<[\vaddr:=\texttt{fpaddr}]> \theta,m) \ast \vaddr \mapsto_{\textsf{vpte}}\; \{\qfrac\} \;\fpaddr \textsf{ wzero 64}}_{\rtv}$
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace} (<[\vaddr:=\texttt{fpaddr}]> \theta,m) \ast \vaddr \mapsto_{\textsf{v}}\; \{\qfrac\} \textsf{wzero 64}}_{\rtv}$

@@

@@

@@

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location of an appropriate L1 table entry,
which is realized in two pieces by the post-condition of \lstinline|ensure_L1|:

@@

\item physical pointsto assertions on the tables L4, L3, and L2 reaching to the
	L1 level entry (l1e) ensure higher levels of the page table exist, 
	are marked present (\textsf{entry\_present} for the relevant
		entries \textsf{l4e l3e l2e}) (Specification Lines 4-6) 
	\item a virtual \emph{PTE}
		pointsto ($\mapsto_{\textsf{vpte}}$) allows access to the memory of the L1 entry
		via virtual address \lstinline|pte_pointsto|.
		A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address is explicit in the assertion (here, \textsf{pa})
		rather than existentially quantified:

@@

@@

	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 

@@

		This supports rules for accessing memory
		at that virtual address, but exposing the physical location being modified
		makes this convenient to use for page table modifications, since we must ensure
		the modified address is the correct physical address that will be used as the L1 entry.
		This is also guaranteed ($ \ulcorner
		\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).

@@

After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
by the physical memory for the L1 entry that will be used to translate the virtual addresses
we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

Then we allocate a fresh page initialized (i.e. zeroed) in Line 14 which return
an \textsf{aligned} address of fresh page (\textsf{fpaddr}) in \textsf{rax}
(Specification Line 19), plus 1 (so the return value is already a valid page table entry).

The crucial point in the proof is then actually updating the L1 entry,
via the virtual address
(\textsf{pte\_addr}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$), to hold the freshly
allocated physical page address (\textsf{fpaddr}) in Line 22.
After this write, we have (Specification Line 26)
\[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by the PTE variant of Rule}~\TirNameStyle{WriteToVirtMemFromReg})\]

At this point in the proof,  $\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr})$ 
contains  full ownerhsip of the L1 table entry physical pointsto $\paddr
\mapsto_{\mathsf{p}} \textsf{fpaddr}$ . Then, we can split the full-ownership of that
physical pointsto
($\textsf{fpaddr}\mapsto_{\textsf{p}} \;\textsf{wzero 64}$) to obtain
fractional ownership on it appropriate for a single address's share of an L1 entry (q4).
Since we know $\texttt{addr\_L1 }(\vaddr,
\entryo) \mapsto_{\mathsf{p}} \{q4\} \;\textsf{fpaddr}$) from
$\mapsto_{\textsf{vpte}}$, after the split of the PTE points-to, we can
connect L4-L2 points-to assertions (in Specification Lines 24 and 25) with 
 $\texttt{fpaddr} \mapsto_{a} (\texttt{wzero 64})$ to obtain the
complete physical table-walk assertion (L4\_L1\_PointsTo(maddr l4e l3e l2e
fpaddr)) in Line 30.

Finally, we can insert $\vaddr$ into the ghost page-table-walk summarization
map ($\theta$) to change its state from unmapped ($\ulcorner \theta
\;!!\;\vaddr = \texttt{None}\urcorner$ in Specification Line 30 and the
precondition) to mapped (Specification Line 33) using Iris' ghost-map update
($\sqsubseteq$), and construct a PTE -points-to $\vaddr
\mapsto_{\textsf{vpte}}\; \qfrac \;\fpaddr (\textsf{ wzero 64})$ (Specification
Line 36) using $\sumwalkabs\vaddr\qfrac\fpaddr$ obtained from ghost
page-table-walk insertions, $\sumapaces\rtv\delta$ from unfolding the
definition of $\mapsto_{\textsf{vpte}}$ in Specification Line 18, and $\fpaddr
\mapsto_{\textsf{p}} \textsf{wzero 64}$ in Specification Line 34.
Finally, by existentially quantifying away \textsf{fpaddr},
we can shift the PTE points-to into a normal virtual points-to.

This proof only maps the first word of the allocated page; the generalization to the full page
is straightforward use of iterated separating conjunction to manage collections of clusters of 512 things
(slices of the L1 entry physical ownership, virtual addresses at 8 byte increments, etc.).

\subsection{Unmapping a Page}
The reverse operation, unmapping a designated page that is currently mapped,
would essentially be the reverse of
the reasoning around line 22 above: given the virtual points-to assertions for all 512
machine words of memory that the L1 entry would map,
and information about the physical location, 
full permission on the L1 entry could be obtained, allowing the construction of a
full virtual PTE pointer for it, setting to 0, and reclaiming the now-unmapped physical memory.

### 2023-04-15 programlogic.tex
stuff

>>>>

\subsection{Soundness}
Our rules from Figure \ref{fig:wpdamd} are proven sound in Iris against an assembly-level hardware model
implementing a fragment of x86-64 including 64-bit address translation with 4-level page tables.

At the moment our proofs do rely on 11 small axioms of properties which should be provable, but
are challenging to discharge due to some representation choices in our model.\footnote{See \lstinline|srx/x64/machine/current_axioms.v|}
